---
title: "Garcia_Acerv_PopStructure"
author: "Sheila Kitchen"
date: "February 13, 2022"
output:
  pdf_document:
    latex_engine: xelatex
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, root.dir="E:/PSU/NOAA/Rocio")
```

```{r warning=FALSE, message=FALSE}
#Required R-packages for multi-locus genotype calling
library(vcfR)
library(poppr)
library(adegenet)
library(ape)
library(ggplot2)
theme_set(theme_bw())
library(knitr)
library(RColorBrewer)
library(data.table)
library(yarrr)
library(reshape2)
library(pcadapt)
library(qvalue)
#library(devtools)
#if (!("OutFLANK" %in% installed.packages())){install_github("whitlock/OutFLANK")}
library(OutFLANK)
library(dplyr)
library(tidyverse)
library(boa)
library("sf")
library("rnaturalearth")
library("rnaturalearthdata")
library(scatterpie)
library(mapplots)

```

#Import the Acerv VCF file:
Based on the genotype report, I extracted the 57 unique MLGs from the 72 Acerv samples using VCFTOOLS (outside of R). I combined those with the unique Acerv MLGs from the STAGdb paper from additional geographic locations (n=121 samples) and filtered for a minor allele frequency of 5% (n= 6,201 SNPs).
```{r import}
##################
### Import VCF ###
##################

#Read in VCF file with array SNVs
ACvcf <- read.vcfR("E:/PSU/NOAA/Rocio/Acerv/AC_mergeDB_maf0.05.recode.vcf")

#add population or species information to the genind pop slot
poptab<-read.table("E:/PSU/NOAA/Rocio/Acerv/acerv_pop3.txt",
                   check.names=FALSE, header=F, na.strings = c("", "NA"),
                   stringsAsFactors = FALSE, sep="\t")
colnames(poptab)<-c("row_id","affy_id","user_specimen_id","region", "lat", "long","subregion", "region2","subregion2","assigned_pop")

gindAC <- vcfR2genlight(ACvcf)
gindAC@pop <- as.factor(poptab$region)
strata(gindAC) <- data.frame(poptab[,c(4,7)])
gindAC

```

# K-clustering- how many clusters/populations?
Identification of the best K is performed by find.clusters in the Adegenet package. This function first transforms the data using PCA. Then, it runs k-means algorithm with increasing values of k and computes associated summary statistics (by default, BIC).
```{r ks}

# find optimal k, repeat 10 times for each value
maxK <- 10
myMat <- matrix(nrow=10, ncol=maxK)
colnames(myMat) <- 1:ncol(myMat)
for(i in 1:nrow(myMat)){
  grp <- find.clusters(gindAC, n.pca = 120, choose.n.clust = FALSE,  max.n.clust = maxK)
  myMat[i,] <- grp$Kstat
}

my_df <- melt(myMat)
colnames(my_df)[1:3] <- c("Group", "K", "BIC")
my_df$K <- as.factor(my_df$K)
head(my_df)

p1 <- ggplot(my_df, aes(x = K, y = BIC))
p1 <- p1 + geom_boxplot()
p1 <- p1 + theme_bw()
p1 <- p1 + xlab("Number of groups (K)")
p1

```
There is a decrease in BIC at k= 4, after which the BIC increases. This suggests 4 clusters should be retained.


# PCA
Summarize the variation between the samples using all the SNP data.
```{r PCA}
ac.pca <- glPca(gindAC, nf = 10)

#plot of the % variation explained by eigenvectors
barplot(100*ac.pca$eig/sum(ac.pca$eig), col = heat.colors(50), main="PCA Eigenvalues")
title(ylab="Percent of variance\nexplained", line = 2)
title(xlab="Eigenvalues", line = 1)

ac.pca.scores <- as.data.frame(ac.pca$scores)
ac.pca.scores$pop <- pop(gindAC)

# variance explained by each axis
var_frac <- ac.pca$eig/sum(ac.pca$eig)

# Colombia regions/subpopulations
Colombia<-c("Urabá Gulf","Cartagena","Rosario", "San Andrés", "San Bernardo", "Tayrona", "I. Arena")

#plot the PCA
set.seed(9)
cols2 <- c(`San Andrés`="#003049",`Urabá Gulf`="#284CB8",`San Bernardo`="#7A94E1",`Rosario`="#3370A3",`I. Arena`="#69BAC7",`Tayrona`="#97CFD8",`Florida`="#959177",`Belize`= "#FCBF49",`Cuba`="#9EDFA8", `Curacao`="#F77F00",`Puerto Rico`="#D62828" ,`USVI`="#961B33")

#cols2 <- c("#FCBF49","#48CAE4","#F77F00","#D62828","#EAE2B7","#E67575","#47A8BD","#ADE8F4","#3370A3","#003049","#8A1919")
p <- ggplot(ac.pca.scores, aes(x=PC1, y=PC2, color=pop))
p <- p + geom_point(size=2)
p <- p + stat_ellipse(level = 0.95, size = 1)
p <- p + scale_color_manual(values=cols2)
p <- p + theme_bw() +ylab("PC2 (6.1%)") + xlab("PC1 (11.4%)")

p

# just Colombia
sub_gindAC<-popsub(gindAC,
  sublist = c("San Andrés","Urabá Gulf","San Bernardo","Rosario","I. Arena","Tayrona")
)

ac.pca <- glPca(sub_gindAC, nf = 10)

#plot of the % variation explained by eigenvectors
barplot(100*ac.pca$eig/sum(ac.pca$eig), col = heat.colors(50), main="PCA Eigenvalues")
title(ylab="Percent of variance\nexplained", line = 2)
title(xlab="Eigenvalues", line = 1)

ac.pca.scores <- as.data.frame(ac.pca$scores)
ac.pca.scores$pop <- pop(sub_gindAC)

# variance explained by each axis
var_frac <- ac.pca$eig/sum(ac.pca$eig)

set.seed(9)
cols2 <- c(`San Andrés`="#003049",`Urabá Gulf`="#284CB8",`San Bernardo`="#69BAC7",`Rosario`="#961B33" ,`Cartagena`="#D62828",`I. Arena`="#F77F00",`Tayrona`="#FCBF49")
p <- ggplot(ac.pca.scores[ac.pca.scores$pop %in% Colombia,], aes(x=PC1, y=PC2, color=pop))
p <- p + geom_point(size=2)
p <- p + stat_ellipse(level = 0.95, size = 1)
p <- p + scale_color_manual(values=cols2)
p <- p + theme_bw() +ylab("PC2 (2.5%)") + xlab("PC1 (14.5%)")

p

```
Capurgana is a transitional zone between San Andres and the other locations. The two samples from Curacao overlap the transitional zone of the Colombia samples.

# DAPC
Multivariate statistical approach that uses populations defined *a priori* to maximize the variance among populations in the sample by partitioning it into between-population and within-population components. This analysis can be used to calculate population membership probabilities.

## Using the optimal K determined above
```{r dapc}
#use best k from above
set.seed(9)
ac.dapc <- dapc(gindAC, grp$grp, n.pca=40, n.da=3)
summary(ac.dapc)

cols<-c("#D62828","#003049","#FCBF49","#EAE2B7")

#scatterplot of dapc
#scatter(ac.dapc, col = cols, cex = 2, legend = TRUE, clabel = F,
#        posi.leg = "bottomleft",posi.da = "topleft", cleg = 0.75,cell=1, cstar=1)

my_df <- as.data.frame(ac.dapc$ind.coord)
my_df$Group <- ac.dapc$grp
my_df$pop <-poptab$region
head(my_df)

p2 <- ggplot(my_df, aes(x = LD1, y = LD2, color = factor(Group), shape=factor(pop)))
p2 <- p2 + geom_point(size = 3, alpha=0.8)
p2 <- p2 + theme_bw()
p2 <- p2 + scale_color_manual(values=cols)+
  scale_shape_manual(values=c(17,18,19,7,21,22,23,24,25,8,15,9))
p2
```

```{r}
#separate by populations
dapc.results <- as.data.frame(ac.dapc$posterior)
dapc.results$pop <- pop(gindAC)
dapc.results$indNames <- rownames(dapc.results)

#reorganize dataframe for ggplot
dapc.results <- melt(dapc.results)
colnames(dapc.results) <-c("Original_Pop","Sample","Assigned_Pop","Posterior_membership_probability")
dapc.results$userID <- poptab$user_specimen_id


dapc.results$Original_Pop = factor(dapc.results$Original_Pop, levels=c( "Florida","Cuba","Belize","San Andrés","Urabá Gulf","San Bernardo","Rosario","I. Arena","Tayrona","Curacao","Puerto Rico","USVI"))

p <- ggplot(dapc.results, aes(x=userID, y=Posterior_membership_probability, fill=Assigned_Pop))
p <- p + geom_bar(stat='identity',color="darkgrey",size=0.0001)
p <- p + facet_grid(~ Original_Pop, scales = "free_x",space = "free_x")
p <- p + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 5), panel.spacing = unit(0, "lines"))+
    scale_y_continuous(expand = c(0, 0))
p <- p + scale_fill_manual(values=cols)
p

```
We can see the two samples from Capurgana were assigned to pop2 and the rest to pop4.

## Map the assigned populations to their geographic locations

```{r maps}
world <- ne_countries(scale = "medium", returnclass = "sf")
class(world)

poptab$assign1<-ac.dapc$assign==1
poptab$assign1[poptab$assign1=="TRUE"]<-1
poptab$assign1[poptab$assign1=="FALSE"]<-0
poptab$assign2<-ac.dapc$assign==2
poptab$assign2[poptab$assign2=="TRUE"]<-1
poptab$assign2[poptab$assign2=="FALSE"]<-0
poptab$assign3<-ac.dapc$assign==3
poptab$assign3[poptab$assign3=="TRUE"]<-1
poptab$assign3[poptab$assign3=="FALSE"]<-0
poptab$assign4<-ac.dapc$assign==4
poptab$assign4[poptab$assign4=="TRUE"]<-1
poptab$assign4[poptab$assign4=="FALSE"]<-0

ag<-aggregate(cbind(assign1,assign2,assign3,assign4) ~ subregion, data = poptab, sum)

ag2 <-ag %>%
  right_join(poptab %>%
      select("subregion","lat", "long","region"),by='subregion') %>%
  group_by(region) %>%
  distinct(subregion,.keep_all = TRUE) %>%
  mutate(radius=sum(assign1,assign2,assign3,assign4)/15) %>%
  mutate(pop1=sum(assign1),pop2=sum(assign2),pop3=sum(assign3),pop4=sum(assign4), n=sum(assign1,assign2,assign3,assign4)) %>%
  distinct(region,.keep_all = TRUE) %>%
  select(-assign1, -assign2,-assign3,-assign4)

p5<-ggplot(data = world) +
    geom_sf() +
    coord_sf(xlim = c(-60, -90), ylim = c(7, 30), expand = FALSE)

p6 <-p5+ geom_scatterpie(data = ag2,aes(x =long, y=lat, group=region, r=radius), cols=c("pop1","pop2","pop3","pop4"),alpha=0.5)+geom_scatterpie_legend(ag2$radius,x=-68, y=25, n=4)
p6 <- p6 + scale_fill_manual(values=cols)
p6

```
### Zoom in on the split assignment by subregions in Capurgana

The collection site Bajo Naui has two samples assigned to pop2 and one sample assigned to pop 4.

```{r map-zoom}

p8<-ggplot(data = world) +
    geom_sf() +
    coord_sf(xlim = c(-77.6, -77.1), ylim = c(8.5, 9), expand = TRUE)

ag3 <-ag %>%
  right_join(poptab %>%
      select("subregion","lat", "long","region"),by='subregion') %>%
  group_by(subregion) %>%
  distinct(subregion,.keep_all = TRUE) %>%
  mutate(radius=sum(assign1,assign2,assign3,assign4)/15) %>%
  mutate(pop1=sum(assign1),pop2=sum(assign2),pop3=sum(assign3),pop4=sum(assign4), n=sum(assign1,assign2,assign3,assign4)) %>%
  select(-assign1, -assign2,-assign3,-assign4)

p9<-p8+ geom_scatterpie(data = ag3,aes(x =long, y=lat, group=subregion, r=radius/10), cols=c("pop1","pop2","pop3","pop4"),alpha=.8)
p9 <- p9 + scale_fill_manual(values=cols)
p9

```
Now we can see where the individuals from Capurgana were collected and their population assignment.

## testing different cluster sizes
We can look for further sub-structure in the data by collection site, although the best K-means cluster is 4.
```{r dapc_multiKs}
my_k <- 2:6

grp_l <- vector(mode = "list", length = length(my_k))
dapc_l <- vector(mode = "list", length = length(my_k))

for(i in 1:length(dapc_l)){
  set.seed(7)
  grp_l[[i]] <- find.clusters(gindAC, n.pca = 40, n.clust = my_k[i])
  dapc_l[[i]] <- dapc(gindAC, pop = grp_l[[i]]$grp, n.pca = 40, n.da = my_k[i])
}

my_df <- as.data.frame(dapc_l[[ length(dapc_l) ]]$ind.coord)
my_df$Group <- dapc_l[[ length(dapc_l) ]]$grp
my_df$pop <-poptab$region
head(my_df)

p2 <- ggplot(my_df, aes(x = LD1, y = LD2, color = factor(pop), shape=factor(Group)))
p2 <- p2 + geom_point(size = 3, alpha=0.8)
p2 <- p2 + theme_bw()
p2 <- p2 + scale_color_manual(values=cols2)+
  scale_shape_manual(values=c(17,18,19,20,21,22,23,24))
p2
```

```{r}
tmp <- as.data.frame(dapc_l[[1]]$posterior)
tmp$K <- my_k[1]
tmp$Genet <- rownames(tmp)
tmp <- melt(tmp, id = c("Genet", "K"))
names(tmp)[3:4] <- c("Group", "Posterior")
tmp$Region <- poptab$region
my_df <- tmp

for(i in 2:length(dapc_l)){
  tmp <- as.data.frame(dapc_l[[i]]$posterior)
  tmp$K <- my_k[i]
  tmp$Genet <- rownames(tmp)
  tmp <- melt(tmp, id = c("Genet", "K"))
  names(tmp)[3:4] <- c("Group", "Posterior")
  tmp$Region <- poptab$region

  my_df <- rbind(my_df, tmp)
}

grp.labs <- paste("K =", my_k)
names(grp.labs) <- my_k
my_df$userID <- poptab$user_specimen_id

my_df$Region = factor(my_df$Region, levels=c("Florida","Cuba","Belize","San Andrés","Urabá Gulf","San Bernardo","Rosario","I. Arena","Tayrona","Curacao","Puerto Rico","USVI"))

cols3 <- c(`1`="#FCBF49",`2`="#003049",`3`="#D62828",`4`="#69BAC7",`5`="#EAE2B7",`6`="#F77F00")
p3 <- ggplot(my_df, aes(x = userID, y = Posterior, fill = Group))
p3 <- p3 + geom_bar(stat = "identity",color="darkgrey",size=0.0001)
p3 <- p3 + facet_grid(K ~ Region, scales = "free_x", space = "free",
                      labeller = labeller(K = grp.labs))
p3 <- p3 + theme_bw()
p3 <- p3 + ylab("Posterior probability of group membership")
p3 <- p3 + theme(legend.position='none')
p3 <- p3 + scale_fill_manual(values=c(cols3))
p3 <- p3 + theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 5), panel.spacing = unit(0, "lines"))+
    scale_y_continuous(expand = c(0, 0))
p3

```
At k=5 Cuba, San Andres and Curacao become a separate population from the eastern pops USVI and Puerto Rico.

# Admixture cross-validation plot
```{r}
#Cross validation plot
library(psych)
ktab<- read.table("E:/PSU/NOAA/Rocio/Acerv/admixture/CVerror_AC.txt", header=F)

tab<-describeBy(ktab$V2, ktab$V1, mat=T)

ggplot(tab, aes(x=as.numeric(group1), y=mean)) +
  geom_errorbar(aes(ymin=mean-sd, ymax=mean+sd), width=0.1) +
  geom_point(size=2)+ xlab("K")+ylab("Prediction Error")+theme_set(theme_bw())+
  scale_x_continuous(breaks = seq(0, 10, by = 1))

```
```{r}
#separate by populations
amix<- read.table("E:/PSU/NOAA/Rocio/Acerv/admixture/admix_k4_clumpak.txt", header=T,check.names=FALSE,stringsAsFactors = FALSE, sep="\t")

#reorganize dataframe for ggplot
admix.results <- melt(amix)

admix.results$Original.pop = factor(admix.results$Original.pop, levels=c("Florida","Cuba","Belize","San Andrés","Urabá Gulf","San Bernardo","Rosario","I. Arena","Tayrona","Curacao","Puerto Rico","USVI"))

p <- ggplot(admix.results, aes(x=sample.id, y=value, fill=variable))
p <- p + geom_bar(stat='identity',color="darkgrey",size=0.0001)
p <- p + facet_grid(~ Original.pop, scales = "free_x",space = "free_x")
p <- p + theme(axis.text.x = element_blank(), axis.ticks.x=element_blank(), panel.spacing = unit(0, "lines"))+
    scale_y_continuous(expand = c(0, 0))
p <- p + scale_fill_manual(values=c(Pop1="#FCBF49",Pop2="#D62828",Pop3="#EAE2B7",Pop4="#003049"))
p

```

# AMOVA
Analysis of Molecular Variance (AMOVA) is a method for estimating population differentiation from molecular data taking into account the mutational distance between alleles. Unlike Fst, which quantifies genetic differentiation based on allele frequencies, AMOVA treats molecular data as vectors and estimates Euclidean distances between alleles. Furthermore, it is possible to test hypotheses about differentiation by grouping subpopulations in a hierarchical structure (Excoffier et al., 1992).
```{r}
gindAC <- vcfR2genind(ACvcf)
gindAC@pop <- as.factor(poptab$region2)
strata(gindAC) <- data.frame(poptab[,c(4,7,8,9,10)])
gindAC

table(strata(gindAC, ~region2/subregion2))

xdis<-bitwise.dist(gindAC)

amova <- poppr.amova(gindAC, ~region2,  dist=xdis)

amova

set.seed(1999)
signif <- randtest(amova, nrepet = 9999)
signif
plot(signif)

amova2 <- poppr.amova(gindAC, ~assigned_pop/region2,  dist=xdis)

amova2

set.seed(1999)
signif2 <- randtest(amova2, nrepet = 9999)
signif2
plot(signif2)
```
There is significant evidence of population structure by region (not surprising).

# FST
```{r}
library(hierfstat)
library("StAMPP")

sf<-stamppFst(gindAC, nboots = 100, percent = 95, nclusters = parallel::detectCores() - 3)

#write.table(sf,"FST_CI_values.txt", sep="\t")

#summary by region
#summary by region
# separate populations into individual genind objects

gindAC <- vcfR2genind(ACvcf)
gindAC@pop <- as.factor(poptab$region)
strata(gindAC) <- data.frame(poptab[,c(4,7,8,9,10)])
gindAC

n.pop <- seppop(gindAC)

bs_I<-basic.stats(n.pop$`I. Arena`)
bs_I$overall

bs_T<-basic.stats(n.pop$Tayrona)
bs_T$overall


bs_SB<-basic.stats(n.pop$`San Bernardo`)
bs_SB$overall

bs_SA<-basic.stats(n.pop$`San Andrés`)
bs_SA$overall

bs_R<-basic.stats(n.pop$Rosario)
bs_R$overall

bs_Cap<-basic.stats(n.pop$`Urabá Gulf`)
bs_Cap$overall

he_ho<-cbind(TI_He=bs_TI$perloc$Hs,TI_HO=bs_TI$perloc$Ho,  SB_He=bs_SB$perloc$Hs, SB_HO=bs_SB$perloc$Ho,
            R_He=bs_R$perloc$Hs,  R_HO=bs_R$perloc$Ho, SA_He=bs_SA$perloc$Hs,SA_HO=bs_SA$perloc$Ho,
            Cap_He=bs_Cap$perloc$Hs, Cap_HO=bs_Cap$perloc$Ho)

boxplot(he_ho)

t.test(bs_TI$perloc$Hs,bs_TI$perloc$Ho, pair = T, var.equal = TRUE, alter = "greater")

# Estimates of heterozygosity for the whole species
sub_gindAC<-popsub(gindAC,
  sublist = c("Urabá Gulf","Rosario","San Andrés","San Bernardo","Tayrona", "I. Arena"),
)
toto_AC <- summary(sub_gindAC)

# plot Apalm and Acerv estimates together
boxplot(toto_AP$Hexp, toto_AP$Hobs,toto_AC$Hexp, toto_AC$Hobs)

# significance test
bartlett.test(list(toto_AC$Hexp, toto_AC$Hobs))
t.test(toto_AC$Hexp, toto_AC$Hobs, pair = T, var.equal = TRUE, alter = "greater")

```
# Nucleotide diverstiy estimate
```{r}

# reload in genlight object
gindAC <- vcfR2genlight(ACvcf)
gindAC@pop <- as.factor(poptab$region)
strata(gindAC) <- data.frame(poptab[,c(4,7)])
gindAC

#source("C:/Users/Sheila's Comp/Downloads/SambaR-master/SambaR-master/SAMBAR_v1.05.txt")

#getpackages(myrepos='http://cran.us.r-project.org',mylib=NULL)

sub_gindAC<-popsub(gindAC,
  sublist = c("Urabá Gulf","Rosario","San Andrés","San Bernardo","Tayrona"))

genlight2sambar(genlight_object="sub_gindAC",do_confirm=TRUE)

filterdata(indmiss=0.25,snpmiss=0.1,min_mac=2,dohefilter=TRUE,min_spacing=500,nchroms=NULL,TsTvfilter=NULL)

findstructure(Kmax=6,add_legend=TRUE,legend_pos="right",legend_cex=2,pop_order=NULL)

calcpi(pi_per_pop=TRUE,myinput=mygenlight,popnames=mysambar$populations,corrected=TRUE)

# part of the seqdivplot() in SambaR
pi_df_AC			<- mysambar$tajdlist_meta[[1]]
	pi_df_AC$nsegsites	<- NA
	nsnps				<- nrow(snps)
	npops				<- length(mysambar$populations)
	mycols				<- mysambar$mycolours[1:npops]
	combipoptable		<- combn(mysambar$populations,m=2)
	combicoltable		<- combn(mycols,m=2)
	combicolvector		<- vector()
	combipopvector		<- vector()
	npairwise			<- ncol(combicoltable)
	ntotal				<- npops+npairwise
	for(i in c(1:npairwise))
		{
		mycol1				<- combicoltable[1,i]
		mycol2				<- combicoltable[2,i]
		mycolfunc3 			<- colorRampPalette(c(mycol1,mycol2))
		combicolvector[i]	<- mycolfunc3(3)[2]
		combipopvector[i]	<- ifelse(combipoptable[1,i]<combipoptable[2,i],paste(combipoptable[1,i],combipoptable[2,i],sep="_"),paste(combipoptable[2,i],combipoptable[1,i],sep="_"))
		}
	for(i in c(1:nrow(pi_df_AC)))
		{
		mypop1				<- pi_df_AC$pop1[i]
		mypop2				<- pi_df_AC$pop2[i]
		mymaf1				<- snps[,paste("maf",mypop1,sep="_")]
		mymaf2				<- snps[,paste("maf",mypop2,sep="_")]
		pi_df_AC$comparison[i]<-ifelse(pi_df_AC$pop1[i]==pi_df_AC$pop2[i],pi_df_AC$pop1[i],ifelse(pi_df_AC$pop1[i]<pi_df_AC$pop2[i],paste(pi_df_AC$pop1[i],pi_df_AC$pop2[i],sep="\n"),paste(pi_df_AC$pop2[i],pi_df_AC$pop1[i],sep="\n")))
		pi_df_AC$comparison2[i]<-ifelse(mypop1==mypop2,"within","between")
		pi_df_AC$nsegsites[i]	<- length(mymaf1[mymaf1>0|mymaf2>0])
	}

# subset to only Colombia samples
pi_df_AC<-pi_df_AC[pi_df_AC$comparison %in% Colombia,]

mymin<- min(pi_df_AC$ndiffpersite,na.rm=TRUE)
mymax<- max(pi_df_AC$ndiffpersite,na.rm=TRUE)

pi_df_AC$comparison <- factor(pi_df_AC$comparison, levels=c("San Andrés","Urabá Gulf","San Bernardo","Rosario","Cartagena","I. Arena","Tayrona"))

boxplot(pi_df_AC$ndiffpersite[pi_df_AC$comparison2=="within"]~pi_df_AC$comparison[pi_df_AC$comparison2=="within"],las=3,ylim=c(mymin,0.57),col=mycols,ylab="",xlab="")

# test for significance between regions
mod<-aov(pi_df_AC$ndiffpersite[pi_df_AC$comparison2=="within"]~pi_df_AC$comparison[pi_df_AC$comparison2=="within"])

summary(mod)

library(multcompView)
TH<-TukeyHSD(mod)
P.value = TH$`pi_df_AC$comparison[pi_df_AC$comparison2 == "within"]`[,4]
P.value = TH$`pi_df$comparison[pi_df$comparison2 == "within"]`[,4]

multcompLetters(P.value)

pi_df %>%
   dplyr::group_by(comparison) %>%
   dplyr::summarize(mean=mean(ndiffpersite), se=sd(ndiffpersite))
```

# Mantel's test
```{r}
library(vegan)
library(ggpmisc)

AC.mat2 <- read.table("geoDist2.txt", header=T,sep="\t",
                               check.names=FALSE,stringsAsFactors = FALSE)

AC.xy <- dist(cbind(AC.mat2$long, AC.mat2$lat))

#Calculate the distance between two locations/nests.
library(geosphere)
#default unit of measurement is kilometers
AC.xy  <- distm(as.matrix(AC.mat2[,c("long","lat")]), fun = distHaversine)/1000

# genetic distance of all sites
genDist<-as.dist(sf$Fsts)

# just Colombia sites
genDist2<-as.dist(sf$Fsts[7:12,7:12])
AC.xy2<-as.dist(AC.xy[7:12,7:12])

AC.correlog <- mantel.correlog(genDist2, XY=AC.xy2, nperm=9999)
summary(AC.correlog)
AC.correlog
plot(AC.correlog)

man<-mantel.rtest(genDist2, as.dist(AC.xy2), nrepet = 9999)
man

aa = as.vector(genDist2)
tt = as.vector(AC.xy2)
mat = data.frame(aa,tt)

my.formula <- y ~ x
mm = ggplot(mat, aes(y = aa, x = tt)) +
    geom_point(size = 3, alpha = 0.5) +
    geom_smooth(method = "lm", colour = "black", alpha = 0.2) +
    labs(x = "Geographic Distance (km)", y = "Genetic Distance (FST)") +
    theme( axis.text.x = element_text(face = "bold",colour = "black", size = 12),
        axis.text.y = element_text(face = "bold", size = 11, colour = "black"),
        axis.title= element_text(face = "bold", size = 14, colour = "black"),
        panel.background = element_blank(),
        panel.border = element_rect(fill = NA, colour = "black"))
mm
```

# Identify SNPs that underlie the split the clusters
## PCAdapt[https://bcm-uga.github.io/pcadapt/articles/pcadapt.html]
This package identifies 584 outlier SNPS associated with PC1 that separate San Andres and the two individuals of Capurgana (population #1) from the other samples based on differences in allele frequencies. The program pcadapt package performs genome scans for selection based on individual genotype data.

```{r pcadapt}
path <- "E:/PSU/NOAA/Rocio/Acerv_uniqueMLG.recode.vcf"

fileAC <- read.pcadapt(path, type = "vcf")

poptab<-read.table("E:/PSU/NOAA/Rocio/Acerv/acerv_pop",
                   check.names=FALSE, header=F, na.strings = c("", "NA"),
                   stringsAsFactors = FALSE, sep="\t")
colnames(poptab)<-c("row_id","affy_id","user_specimen_id","region", "lat", "long","subregion","assigned_pop")

# run pcadapt, test up to 20 clusters/pops
acp <- pcadapt(input = fileAC, K = 20)

# Screeplot displays in decreasing order the percentage of variance explained by each PC
plot(acp, option = "screeplot",  K = 20)
```
It was found that K=5 corresponds to the optimal choice of the number of PCs.

```{r pcadapt_2}
#displays the PCA results colored by collection regions
plot(acp, option = "scores", pop = poptab$assigned_pop, gg.col=transp(cols,0.6))

#change to different PC axes
plot(acp, option = "scores", i = 1, j = 3, pop = poptab$assigned_pop, gg.col=transp(cols,0.6))
plot(acp, option = "scores", i = 1, j = 4, pop = poptab$assigned_pop, gg.col=transp(cols,0.6))

plot(acp, option = "scores", i = 2, j = 3, pop = poptab$assigned_pop, gg.col=transp(cols,0.6))
plot(acp, option = "scores", i = 2, j = 4, pop = poptab$assigned_pop, gg.col=transp(cols,0.6))
```

Looking at population structure beyond K = 2 confirms the results of the scree plot. The third and the fourth principal components do not ascertain population structure anymore.


Re-run with the best PC based on the screen plot above but doing genome scan for each principal component to get component-wise p-values. The test statistics are the loadings, which correspond to the correlations between each PC and each SNP. P-values are computed by making a Gaussian approximation for each PC and by estimating the standard deviation of the null distribution.

```{r pcadapt_3}
y <- pcadapt(fileAC, K = 2, method = "componentwise",LD.clumping = list(size = 200, thr = 0.2),min.maf = 0.05) #K is the best number of PC from above
summary(y)
summary(y$pvalues)
```
A Manhattan plot displays −log10 of the p-values.
```{r pcadapt_4}
plot(y , option = "manhattan", K=2)
```

Check the expected uniform distribution of the p-values using a Q-Q plot and histogram of the pvalues to check the distribution.
```{r pcadapt_5}
plot(y, option = "qqplot", threshold = 0.1,K=2)

hist(y$pvalues, xlab = "p-values", main = NULL, breaks = 50, col = "orange")
plot(y, option = "stat.distribution",K=2)
```

The p-values were transformed into qvalues to control for false discovery rate, or the percentage of false discoveries among the list of candidate SNPs. I used a false discovery rate lower than 1% (alpha = 0.01).
```{r pcadapt_6}
qval1 <- qvalue(y$pvalues[,1])$qvalues
alpha <- 0.01
outliers1 <- which(qval1 < alpha)
length(outliers1)

snp_pc1 <- get.pc(y, outliers1)
# PC1
length(which(snp_pc1$PC =="1"))
#PC2
length(which(snp_pc1$PC =="2"))

```
## OutFLANK[https://rpubs.com/lotterhos/outflank]
Fst measures the degree of allele frequency differences between 2 or more groups of individuals. In most cases, these groups represent different spatially distinct populations. However, population demographic history (drift, bottlenecks, range expansions) can have a profound effect on the distribution of Fst, and often leads to increases in the number of high Fst loci, leading to false positives in scans for selective sweeps. Therefore, we need to be careful to separate out loci with high Fst due to neutral demographic history from those with high Fst due to natural selection.

OutFLANK is an R package that implements the method developed by Whitlock and Lotterhos to use likelihood on a trimmed distribution of FST values to infer the distribution of FST for neutral markers. This distribution is then used to assign q-values to each locus to detect outliers that may be due to spatially heterogeneous selection.

Conversion of the vcf into the correct format for this package:
```{r outflank}
#convert vcf to OUTFLANK format
ACvcf <- read.vcfR("E:/PSU/NOAA/Rocio/Acerv_uniqueMLG.recode.vcf")
geno <- extract.gt(ACvcf)
G <- matrix(NA, nrow = nrow(geno), ncol = ncol(geno))

G[geno %in% c("0/0", "0|0")] <- 0
G[geno %in% c("0/1", "1/0", "1|0", "0|1")] <- 1
G[geno %in% c("1/1", "1|1")] <- 2

sum(is.na(G)) # should not contain NAs
o2 <- G
o2[is.na(o2)] <- 9
sum(is.na(o2))
dim(o2)
levels(factor(as.numeric(unlist(o2))))

table(as.vector(o2))
SNPdata <- t(o2)
head(SNPdata[,1:20])

#pull out chrom and position from VCF file
t <- getCHROM(ACvcf)
u <- getPOS(ACvcf)
r<- paste(t,u,sep="_") #loci names
```

A FST is calculated for each SNP based on the orignal collection site (regions):
```{r outflank_2}
FstDataFrame <- MakeDiploidFSTMat(SNPdata, r, poptab$region)

head(FstDataFrame)

plot(FstDataFrame$FST, FstDataFrame$FSTNoCorr, xlim = c(-0.01,0.3),
     ylim = c(-0.01, 0.3), pch = 20)
abline(0, 1) # Checking the effect of sample size on Fst since FSTCoCorr will be used in the follow

plot(FstDataFrame$He, FstDataFrame$FSTNoCorr, pch=20, col="grey")

hist(FstDataFrame$FSTNoCorr, breaks=seq(0,1, by=0.001))
hist(FstDataFrame$FSTNoCorr[FstDataFrame$He>0.05], breaks=seq(0,1, by=0.001))
hist(FstDataFrame$FSTNoCorr[FstDataFrame$He>0.1], breaks=seq(0,1, by=0.001))
```

Next, trim back loci in the upper and lower 5% of the empirical Fst distribution, and eliminate low frequency SNPs (ones with expected heterozygosity, Hmin, < 0.1; equates to a MAF of ~0.05).

```{r outflank_3}
OF <- OutFLANK(FstDataFrame, NumberOfSamples=2, qthreshold = 0.05,
               RightTrimFraction = 0.05,LeftTrimFraction = 0.25)

# Plot the ditribution of Fst with the chi squared distribution, the empirical distribution of Fst (yellow bars) and the predicted neutral distribution (black line)
OutFLANKResultsPlotter(OF, withOutliers = TRUE, NoCorr = TRUE, Hmin = 0.1,
                       binwidth = 0.005, Zoom = FALSE, RightZoomFraction = 0.05,
                       titletext = NULL)

OutFLANKResultsPlotter(OF , withOutliers = TRUE,
                       NoCorr = TRUE, Hmin = 0.1, binwidth = 0.005, Zoom =
                         TRUE, RightZoomFraction = 0.15, titletext = NULL)

hist(OF$results$pvaluesRightTail)

sum(OF$results$qvalues<0.05, na.rm=TRUE)
```

Now detect outlier high FST markers.
```{r outflank_4}
P1 <- pOutlierFinderChiSqNoCorr(FstDataFrame, Fstbar = OF$FSTNoCorrbar,
dfInferred = OF$dfInferred, qthreshold = 0.05, Hmin=0.1)
head(P1)
tail(P1)

my_out <- P1$OutlierFlag=="TRUE"
plot(P1$He, P1$FST, pch=19, col=rgb(0,0,0,0.1), ylim=c(-0.02,1))
points(P1$He[my_out], P1$FST[my_out], col="blue")
```
This results in 0 outlier SNPs.

## BayeScan
Yet another tool to find high-scoring FST outliers from the SNP data. BayeScan aims at identifying candidate loci under natural selection from genetic data, using differences in allele frequencies between populations.The difference in allele frequency between this common gene pool and each subpopulation is measured by a subpopulation specific FST coefficient.BayeScan incorporates the uncertainty on allele frequencies due to small sample sizes because it is Bayesian method.

Export the genotype matrix for use in Bayescan outside the R environment.
```{r Bayescan}
#convert format of SNPdata for bayescan
rownames(SNPdata)<-t(colnames(geno))
#write.table(SNPdata, "acerv_geno_snps.txt", sep=" ")
```

Import the result of Bayescand the R script from Bayescan to plot the results and find the outliers.
```{r Bayescan_2}
source("C:/Users/Sheila's Comp/Documents/BayeScan2.1/BayeScan2.1/R functions/plot_R.r")

fst <- plot_bayescan("E:/PSU/NOAA/Rocio/BayeScan/ac_aug19_fst.txt", FDR=0.05, add_text=F)
summary(fst)
#fst$outliers
fst$nb_outliers

b <- read.table("E:/PSU/NOAA/Rocio/Bayescan/ac_aug19_fst.txt")
b$ID <- seq.int(nrow(b))
colnames(b) <- c("BAYES_prob", "BAYES_log10.PO.", "BAYES_qval", "BAYES_alpha", "BAYES_fst", "ID")

fout <- as.data.frame(fst$outliers)
colnames(fout) <- c("ID")

sel <- read.table("E:/PSU/NOAA/Rocio/Bayescan/ac_aug19.sel", colClasses = "numeric")
parameter="Fst5"
plot(density(sel[[parameter]]), xlab=parameter, main=paste(parameter, "posterior distribution"))
boa.hpd(sel[[parameter]], 0.05)
```
This found 260 outlier SNPs.

## Combining the three outlier detection methods

```{r combine reports}
# make a data frame of q/pvalues
qvalz <- as.data.frame(qval1)
pvalz <- as.data.frame(y$pvalues)
out2<-cbind(as.integer(rownames(pvalz)),r, pvalz,qvalz)
head(out2)

colnames(out2)<-c("ID", "LocusName","pca_pvalue_p1","pca_pvalue_p2", "pca_qvalue")

outliers<- out2  %>%
  mutate(PCA_outlier = ifelse(pca_qvalue < 0.01, "TRUE", "FALSE")) %>%
  mutate(PCA_pc1 = ifelse(out2$ID %in% snp_pc1$SNP, snp_pc1$PC, "FALSE")) %>%
  left_join(b %>% select("BAYES_qval","BAYES_fst", "ID"), by=c("ID")) %>%
  mutate(BAYES_outlier = ifelse(ID %in% fout$ID, "TRUE", "FALSE")) %>%
separate(LocusName, into = c("chrom", "pos"), sep="_(?=[^_]+$)")

#write.table(outliers, file="outliers_PC1_AC_update.txt", sep="\t")
```

Create a venn diagram of what outliers SNPs are shared from each method.
```{r venndiagram}
library(VennDiagram)
#1
nrow(subset(outliers, PCA_pc1== "1"))
#2
nrow(subset(outliers, BAYES_outlier== "TRUE"))
#1 and 2
nrow(subset(outliers,  PCA_pc1== "1" & BAYES_outlier== "TRUE"))

pdf(file="AC_venn.pdf",width=5, height=5)
grid.newpage()
draw.pairwise.venn(area1 = 584, area2 = 260, cross.area = 165, category = c("PCAdapt", "BayeScan"), lty = "blank",
                 fill = c("#0C5BB0FF", "#EE0011FF"), fontfamily =
                   rep("sans", 7), cat.fontfamily = rep("sans", 3))
dev.off()
```
